---
title: "Longitudinal effects of working memory on L2 grammar and reading abilities"
subtitle: "Sagarra (2017)"
author: "Kaylee Fernandez"
institute: "Rutgers University"
date: "April, 7, 2021"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

```{r setup, echo = FALSE, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(scipen = 999)
library(tidyverse)
library(knitr)
library(kableExtra)
library(xaringan)
library(patchwork)
library(xaringanthemer)
library(showtext)
``` 

#Introduction
Sagarra (2017) investigated the effects of working memory (WM) on L2 grammar and reading development in adult beginning classroom learners. 

In **Experiment 1**, 82 beginning learners of Spanish completed a WM test without a demanding processing task (Daneman and Carpenter's test) and grammar and reading pretests and posttests one year apart. 

In **Experiment 2**, 330 beginning learners of Spanish completed the same tests as in Experiment 1 and a working memory test without a demanding processing task (Waters and Caplan's test).

---
# Background: Working memory 

While much research has investigated WM and second language acquisition, questions remain regarding whether WM predicts SLA longitudinally.

The most widely used index of WM is reading span 

+ For example, Daneman and Carter's (1980) test asks participants to read sets of sentences aloud at their own pace while remembering the final word of each sentence in the set

Sagarra (2017) suggested that this test is problematic because its processing task is not sufficiently cognitively complex

+ Waters and Caplan (1996) modified Daneman and Carter's test in order to make it more cognitively demanding by asking participants to make plausibility judgments after each sentence under time pressure

+ Three scores are generated by making this change: recall span, logicality judgment, and response time 

---
#Research Questions

In order to explore whether WM predicts SLA longitudinally and whether a cognitively demanding WM test better predicts SLA longitudinally than a less demanding WM test, Sagarra (2017) asked the following research questions: 

+ RQ1: Does WM impact L2 grammar learning and reading compression longitudinally in beginning classroom learners based on recall span of a WM test without a demanding processing task (Daneman and Carter's test)? 

+ RQ2: Does WM impact L2 grammar learning and reading compression longitudinally in beginning classroom learners based on recall span of a WM test with a demanding processing task (Walter and Caplan's test)? 

+ RQ3: Does response time predict grammar and reading development? 

+ RQ4: Are processing (response time) and storage (recall span) associated? (i.e., is there a processing-storage trade-off?)
---
#Hypotheses

Based on research revealing longitudinal WM effects on L2 grammar learning and L2 reading development with WM tests with a demanding processing task, Sagarra (2017) hypothesized that measures of recall span and response time on a WM test with a demanding processing task (i.e., Walter and Caplan's test) would predict grammar and reading development in beginning classroom readers. 
---
# Methods (Experiment 1)
Participants were L1 English learners of Spanish, enrolled in a second semester (pretests) of Spanish and, later, students enrolled in a fourth semester of Spanish. 

The initial sample pool consisted of 382 participants but due to attrition the final sample pool was reduced by 82 participants. 

Participants completed: 
+ Screening tests: A Spanish proficiency test (The university's Spanish placement test) and a language background questionnaire 
+ A working memory test: Daneman and Carpenter's test (Contained 60 unrelated sentences, 13 to 16 words in length)
+ A grammar test (pretest, posttest): A sentence level (Mecartty, 1993) and paragraph level (University of Iowa's Spanish placement test) measuring Spanish grammatical knowledge (The sentence level test contained 12 sentences and the paragraph level test, two cloze tests with with 10 total missing words)
+ A reading test (pretest, posttest): Taken from Mecartyy's dissertation (9 multiple choice questions evaluating reading comprension)
---
#Explanation of analysis 
In Experiment 1, the data were submitted to two types of statistical analyses: **GLM** and **bivariate correlational analysis**.

In regards to the GLM, the author stated that "two repeated-measure ANCOVAs were conducted." (p. 350)

Analysis of covariance (ANCOVA) is used to test the main and interaction effects of categorical variables on a continuous dependent variable, controlling for the effects of other continuous confounding variables, which co-vary with the dependent variable (i.e., covariates). 

A repeated measures ANCOVA compares means across one or more variables that are based on repeated observations while controlling for a confounding variable. 

A repeated measures ANCOVA has at least one dependent variable and one covariate, with the dependent variable containing more than one observation.

---
#Explanation of analysis 
As stated earlier, in Experiment 1, the author reported that two repeated measures ANCOVA's were conducted with Time as a within-participant independent variable (Time 1 = pretest; Time 2 = posttest), and WM as a covariate: one for the grammar pretest and posttest scores, and one for the reading pretest and posttest scores.

A bivariate correlational analysis was used to analyze the relationship between: 
+ WM and grammar pretest and posttest scores
+ WM  reading pretest and reading posttest scores.

Had this been a traditional ANCOVA , it would have tested the effects of a categorical variable (Time) on a continuous dependent variable (grammar or reading scores) based on repeated observations, controlling for the effect of a continuous confounding variable (WM).

However, it would not have been able to test for the interaction between the categorical independent variable (Time) and the continous confounding variable (WM). 

---
#Appropriateness of analysis 
 
As just mentioned, in Sagarra (2017) the categorical independent variable was Time and the continuous covariate was WM. While a mixed GLM can look at the interaction between categorical and continuous predictors using dummy coding, a traditional ANCOVA cannot. 

However, in order to answer the research questions (E.g., *RQ1: Does WM impact L2 grammar learning and reading compression longitudinally in beginning classroom learners based on recall span of a WM test without a demanding processing tast (Daneman and Carter's test)?*) it is necessary to look at the interaction between Time and WM. 

The results (and the research questions) suggest that in fact a true ANCOVA was not run, but rather a mixed GLM.  Identifying the GLM as an ANCOVA may be misleading and, rather than labeling it as such, the model should have been solely identified as a GLM to avoid confusion. 

---

#Presentation of results 
The ANCOVAs showed a significant main effect for Time in the grammar test, *F*(1,80)= 8.105, *p* = .006, but not in the reading test, *F*(1,77)= 4.466, *p* = .038. 
A Bonferonni post hoc test revealed that, in the grammar test, participants scored higher in the posttest than the presttest. There was no significant main effect for WM in the grammar test, *F*(1,80)= .124, *p* = .726, or the reading test, *F*(1,77)= .038, *p* = .846. 
There was no significant interaction of Time and WM in the grammar test, *F*(1,80)= 1.169, *p* = .283, or the reading test, *F*(1,77)= .067, *p* = .796. 
Bivariate correlations showed that WM did not correlate with the grammar pretest (*r* = 0.13, *p* = .902), the grammar posttest (*r* = -.076, *p* = .477), the reading pretest  (*r* = .002, *p* = .988), or the reading posttest (*r* = 0.033, *p* = .768). 

The results shed light on the nature of the GLM, it is clear a traditional ANCOVA was not run although it was specified as such. 

---

#Presentation of results 

The presentation of results could have also been strengthened by including a discussion of the assumptions of the model. For example: *Did visual inspection of Q-Q plots and plot of residuals against fitted vales reveal no discernible deviations from normality and homoscedasticity?*

No tables or graphs were used. Including a table of the descriptive statics, for example, could have facilitated the readability. 

The author properly interpreted the findings, stating that the results suggest that classroom instruction improved L2 grammar learning and reading abilities over the course of two semesters, but that WM did not modulate the improvement. 

Moreover, the author gave several explanations (e.g., methodological limitations of Daneman and Carpenter's test) as to why WM did not modulate the improvement and clearly explained her rationale. 

---
# Methods (Experiment 2)
Participants were L1 English learners of Spanish, enrolled in a third semester of Spanish 

330 participants 

Participants completed: 
+ Screening tests: A Spanish proficiency test (**different from Experiment 1 because the participants were enrolled in different universities**) and a language background questionnaire 
+ **Two working memory tests**: Daneman and Carpenter's test (Contained 60 unrelated sentences, 13 to 16 words in length) and **Waters and Caplan's test** (80 sentences)
+ A grammar test (pretest, posttest): A sentence level (Mecartty, 1993) and paragraph level (University of Iowa's Spanish placement test) measuring Spanish grammatical knowledge (The sentence level test contained 12 sentences and the paragraph level test two cloze tests with with 10 total missing words)
+ A reading test (pretest, posttest): taken from Mecartyy's dissertation (9 multiple choice questions evaluating reading comprension)
---
#Explanation of analysis 

In Experiment 2, the author reported that a series of bivariate correlations and four repeated-measures ANCOVAs were conducted. 

As in Experiment 1, it was stated that in Experiment 2 the four ANCOVAS were conducted with Time as a within-participant independent variable, and WM as a covariate: two for Daneman and Carpenter's test (one for the grammar test, one for the reading test), and two for Waters and Caplan's test (one for the grammar prettest and posstest scores, one for the reading prettest and posstest scores). 

To address RQ3 (whether response time predicts grammar and reading development), the author reported that two additional ANCOVAs were carried out with response time.

To address RQ4, correlations were conducted between recall span and response time scores. 
---
#Appropriateness of analysis 

Just like for Experiment 1, the results of Experiment 2 (and the research questions) suggest that in fact a true ANCOVA was not run, but rather a mixed GLM. 

Since a mixed GLM can look at the interaction between continuous and categorical independent variables, it is the best analysis to answer the research questions.

As stated earlier, the model should not have been identified as an ANCOVA to avoid confusion. 
---
#Presentation of results 

In Experiment 2, the results were presented similarly as to how they were presented in Experiment 1 (e.g., no mention of assumptions, no tables or graphs). 

One notable difference, in addition to reporting the F value, degrees of freedom, and the p value, in Experiment 2 effect site (partial Eta squared: *η2*) was also reported. 

For example, "There was no significant interaction of Time and WM (Daneman and Carpenter's test) in the grammar test, *F*(1,328) = 2.145, *p* = .144, partial *η2* = .006 or the reading test, *F*(1,328) = .841, *p* = .360, partial *η2* = .002." (p. 354)

---
#Presentation of results 

No explicit explanation was given for why *η2* (a measure of effect size which reflects the percentage of the variance in the dependent variable explained by the independent variables in a sample) was reported in Experiment 2 and not in Experiment 1. 

However, it could be understood implicitly that *η2* was reported in order to examine whether p values were significant due to having a large sample pool. 

For example, the author states, "Importantly, the partial *η2* values show a small effect size, meaning that the significant *p* value was not due ot having a large sample pool." (p. 354)

Otherwise, the author accurately explained the results and clearly explained the rationale behind her interpretation of the findings. 
---
#Pros and Cons of analysis (Experiments 1 and 2)

Cons: 

+ Labeling the model as an ANCOVA is potentially misleading.
+ The assumptions of the model were not reported. Ensuring that the assumptions are met is an important step of data analysis and it should not be assumed. 
+ A more explicit rationale for why *η2* was reported in Experiment 2 but not in Experiment 1 would have been helpful. 

Pros: 
+ What was measured was theoretically motivated. For example, a measure of response time was included based on research confirming the the importance of timing measures to predict working memory span and offer insight into WM functioning
+ While it was not clearly stated, an appropriate model (GLM) for answering the research questions was used. 
+ A clear rationale for the interpretation of the results was provided. 

---
#References

Daneman, M, & Carpenter, P. A. (1980). Individual differences in working memory and reading. *Journal of Verbal Learning and Verbal Behavior, 19*, 450–66.

Sagarra, N. (2017). Longitudinal effects of working memory on L2 grammar and reading abilities. *Second Language Research, 33*(3), 341–363.

Waters, G. S., & Caplan, D (1996). The measurement of verbal working memory capacity and its relation to reading comprehension. *The Quarterly Journal of Experimental Psychology, 49*, 51–79.
